## How important is it to ask pointed questions? {.smaller}

The following is example is taken from [Chapter 18 of Introduction to Modern Statistics](https://openintro-ims.netlify.app/inference-tables.html)

::: vertical-space
:::

In this experiment, 219 individuals were asked to be a seller of an iPod
The participant received $10 + 5% of the sale price for participating. 
The iPod they were selling had frozen twice in the past inexplicably but otherwise worked fine.

::: vertical-space
:::

The prospective buyer starts off and then asks one of three final questions, depending on the seller's treatment group.

- **General:** What can you tell me about it ?
- **Positive Assumption:** It doesn't have any problems, does it ?
- **Negative Assumption:** What problem does it have ?



The outcome variable is whether or not the participant discloses or hides the problem with the iPod


:::footer
[*Source: Minson JA, Ruedy NE, Schweitzer ME. There is such a thing as a stupid question: Question disclosure in strategic communication*](https://www.acrwebsite.org/volumes/1012889/volumes/v40/NA-40)
:::

##

The following data is the `ask` dataset from the R package `openintro`

```{r}
#| echo: true
ask_two_way <- openintro::ask |> 
  select(question_class, response) |> 
  table() |> 
  addmargins()
```



```{r}
attr(ask_two_way,'dimnames')$question_class <- c('General','Negative_assumption','Positive_assumption','Total')
attr(ask_two_way,'dimnames')$response <- c('Disclose','Hide','Total')
knitr::kable(ask_two_way)
```



##

```{r}
#| fig-align: center
openintro::ask |> 
  ggplot() +
  aes(x = question_class, fill = response) +
  geom_bar(position = "fill") +
  scale_fill_manual(values=c('#8dc4b0','#948DC4')) +
  theme(axis.text.x = element_text(angle = 45,vjust = .5)) +
  labs(x = "Question Class",
       y = "Proportion")
```


If there were no relationship, we would expect to see the purple bars reaching to the same height, regardless of the question class



##

:::callout-note
## Expected Counts
To calculate the expected count for the $i^{th}$ row and $j^{th}$ column, compute
$$
\text{Expected Count}_{\,row \, i,\, \;col \, j } = 
\frac{
(\text{row}\,\; i \; \text{total}) \times (\text{column}\,\; j \; \text{total}) 
}{
\text{table total}
}
$$
:::



##
:::callout-note
## $\chi^2$ test-statistic
The test statistic for assessing the independence between two categorical variables

$$
\chi^2 = \sum_{ij} \frac{
(\text{observed count - expected count})^2
}{
\text{expected count}
}
$$
When the null hypothesis is true and the conditions are met, $\, \chi^2$ has a Chi-squared distribution with 
$\, df = (r-1) \times (c-1)$, where $r=$ number of rows, and $c=$ number of columns
:::

Conditions:

- Independent observations
- Large samples: At least 5 expected counts in each cell