## Goodness of Fit: Using `Infer`

Considering a goodness of fit test, we want to look at the responses from the participants in their disclosure/hiding of the problem with the iPods

Suppose our null hypothesis is that `response` follows a uniform distribution *(i.e., there's actually an equal number of hidden and disclosed responses to the problems with the iPods)*


##

```{r}
#| fig-align: center
openintro::ask |>
  count(response) |> 
  mutate(proportion = n/sum(n)) |> 
  ggplot() +
  aes(x = response, y= proportion) +
  geom_bar(stat='identity',fill = '#8dc4b0') +
  geom_hline(yintercept = 0.5,col='red', linetype = 2)+
  labs(x = "response: Disclosure/hiding of the problem with the iPod",
       y = "Proportion of Responses")
```


It seems like there are more hidden responses compared to disclosing the actual problems with the iPods. Lets now test whethere this difference is statistically significant 


##

First, to carry out this hypothesis test, we need to calculate out observed statistic



##

The `infer` package provides a wrapper function `chisq_test`, to carry out $\chi^2$ goodness of fit tests on `tidy data`

```{r}
#| echo: true
chisq_test(openintro::ask, 
           response = response,
           p = c('hide' = 1/2,
                 'disclose' = 1/2))
```



