
## Part 3:  $\chi^2$ Goodness of Fit

The $\chi^2$ goodness of fit test is a statistical test used to determine whether observed categorical data follows an expected distribution or conforms to a specified theoretical distribution. 

::: vertical-space
:::

It assesses whether there is a significant difference between the observed frequencies and the expected frequencies for different categories or group



## Gregor Mendel’s Experiments

From (1822-1884) Gregor Mendel conducted crossing experiments with pea plants
of different shapes and color. Let us look at the outcome of a pea crossing experiment with the following results:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
crossing_peas <- tibble::tibble(cross_result = 
                                  rep(c('round yellow', 'round green',
                                        'edged yellow','edged green'),
                                      c(315,108,101,32) )) |> 
  mutate(cross_result = as.factor(cross_result))
```


::: vertical-space
:::


```{r}
knitr::kable(crossing_peas |> table())
```



## {.smaller}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
#| fig-align: center

positions <- c('round yellow', 'round green','edged yellow','edged green')

crossing_peas |>
  count(cross_result) |> 
  mutate(proportion = n/sum(n)) |> 
  ggplot() +
  aes(x = cross_result, y= proportion) +
  geom_bar(stat='identity',fill = '#8dc4b0') +
  scale_x_discrete(limits = positions)+
  labs(x = "Crossing result",
       y = "Proportion")
```


Mendel had the hypothesis that the four different types occur in proportions of 9:3:3:1, that is
$$
p_{\text{round-yellow}} = \frac{9}{16}, \,
p_{\text{round-green}} = \frac{3}{16}, \,
p_{\text{edged-yellow}} = \frac{3}{16}, \,
p_{\text{edged-green}} = \frac{1}{16},
$$


## {.smaller}

Mendel's hypothesis is then 

$$
H_0: \,
p_{\text{round-yellow}} = \frac{9}{16}, \,
p_{\text{round-green}} = \frac{3}{16}, \,
p_{\text{edged-yellow}} = \frac{3}{16}, \,
p_{\text{edged-green}} = \frac{1}{16},
$$

$$
H_a: \
p_{\text{round-yellow}} \neq \frac{9}{16}, \,
p_{\text{round-green}} \neq \frac{3}{16}, \,
p_{\text{edged-yellow}} \neq \frac{3}{16}, \,
p_{\text{edged-green}} \neq \frac{1}{16},
$$

::: vertical-space
:::


The alternative simply states the proportions of the four different outcomes of a pea crossing experiment is not 9:3:3:1




## Goodness of Fit: Using `Infer`



First, to carry out this hypothesis test, we need to calculate out observed statistic

```{r}
#| echo: true
observed_gof_statistic <- crossing_peas |> 
  specify(response = cross_result) |> 
  hypothesize(null = "point",
              p = c("round yellow" = 9/16,
                    "round green" = 3/16,
                    "edged yellow" = 3/16,
                    "edged green" = 1/16)) |> 
  calculate(stat = "Chisq")
```

```{r}
observed_gof_statistic
```


Now we want to compare the observed $\chi^2 =0.47$ statistic to a null 
distribution, generated under the assumption that the proportions of the four different outcomes of a pea crossing experiment is 9:3:3:1

 

##

::: {.panel-tabset}


### Code

First, generate the null distribution using simulation

```{r}
#| echo: true
null_dist_gof <- crossing_peas |> 
  specify(response = cross_result) |> 
  hypothesize(null = "point",
              p = c("round yellow" = 9/16,
                    "round green" = 3/16,
                    "edged yellow" = 3/16,
                    "edged green" = 1/16)) |> 
  generate(reps = 1000, type = "draw") |> 
  calculate(stat = "Chisq")
```

::: vertical-space
:::


To get a sense for what these distributions look like, and where our observed statistic falls, we can use `visualize()`


```{r}
#| echo: true
#| eval: false
null_dist_gof |> 
  visualize() + 
  shade_p_value(observed_gof_statistic,
                direction = "greater")
```


### Output


```{r}
#| fig-width: 9
#| fig-height: 4
null_dist_gof |> 
  visualize() + 
  shade_p_value(observed_gof_statistic,
                direction = "greater")
```


The statistic seems like it would be quite likely if the proportions of the four different outcomes of a pea crossing experiment is 9:3:3:1. 



:::




## 

Using the simulated null distribution, we can approximate the $\,p\,$-value with `get_p_value`


```{r}
p_value_gof <- null_dist_gof |> 
  get_p_value(observed_gof_statistic,
              direction = 'greater')
```

```{r}
p_value_gof
```

::: vertical-space
:::

Thus, if the proportions of the four different outcomes of a pea crossing experiment is  9:3:3:1, our approximation of the probability that we would see a distribution like the one we did is approximately `r p_value_gof`


::: vertical-space
:::

To calculate the $p$-value using the true $\chi^2$ distribution, we can use the `pchisq` function from base R


```{r}
#| echo: true
pchisq(observed_gof_statistic$stat, df = 3, lower.tail = FALSE)
```




##

The `infer` package provides a wrapper function `chisq_test`, to carry out $\chi^2$ goodness of fit tests on `tidy data`

```{r}
#| echo: true
chisq_test(crossing_peas, 
           response = cross_result, 
           p = c("edged green" = 1/16, # <1>
                 "edged yellow" = 3/16,
                 "round green" = 3/16,
                 "round yellow" = 9/16))
```

1. If `response` is a factor, then list the proportions of the levels in the order of `levels(response)`




::: vertical-space
:::


Thus, we do not have sufficient evidence to suggest Mendel's hypothesis that the four different types occur in proportions of 9:3:3:1
is not wrong. That is we fail to reject the null hypothesis.


*(This does not mean we accept Mendel's hypothesis)*







