---
title: "Linear Regression"
author: "Created Dr. Mine Dogucu and Presented by Dr. Jaynes"
format: 
  revealjs:
    slide-number: true
    logo: "https://socaldatascience.github.io/bootcamp-materials-2022/img/socalds-logo.png"
    theme: ["slide-style.scss"]
    incremental: false
---

## Overview (To be removed)
- 2 hours of lecture (1 hour each)
- 50 minutes of activity (25 mins each)
- 2 5 min breaks

## Goals
- Review Linear Regression Concepts from your Data Science Courses 
- Conditions for Least Squares Regression and Diagnostics
- Multiple Linear Regression
- Model Evaluation 

## Linear Regression Activity
- Simple Linear Regression: 1) Numerical predictor, 2) categorical predictor
- Scatterplot
- Fit model in R
- Model output
- Model equation
- Interpretation of estimates
- Prediction
<!--
https://mdsr-book.github.io/mdsr2e/ch-foundations.html#sec:gooddesign
-->

## Linear Regression Review
- Scatterplot
- Response and explanatory variables
- Linear equation: Math vs stats class
- Using R to fit a linear model
- Equation of linear model
- Interpretation of estimates
- Extrapolation
- Prediction
- Residuals
- Least squares regression
<!--
https://r4ds.had.co.nz/model-basics.html#model-basics
-->

## Conditions for Least Squares Regression
<!--
https://openintro-ims.netlify.app/inf-model-slr.html#tech-cond-linmod
-->

## Multiple Linear Regression
- Numerical and categorical predictor (binary and multi-category)
- Interpretation
- Prediction

## Model Evaluation
- R2
- RMSE
- Overfitting
- Splitting the data


```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(openintro)
library(broom)
theme_set(theme_gray(base_size = 18))
```





<!--

## Data `babies` in `openintro` package

```{r echo = FALSE}


glimpse(babies)

```

##  Baby Weights
:::: {.columns}
::: {.column width="40%"}
```{r eval = FALSE, fig.align='center', message = FALSE, warning = FALSE}
ggplot(babies, 
       aes(x = gestation, y = bwt)) +
  geom_point()

```
:::

::: {.column width="60%"}
```{r echo = FALSE, fig.align='center', message = FALSE, fig.height= 6,warning = FALSE}


ggplot(babies, 
       aes(x = gestation, y = bwt)) +
  geom_point()

```
:::
::::

## Response and Explanatory Variables

| Variable    | Variable Name | Type of Variable |
|-------------|-----------------|---------|
| Response (y)    | Birth weight | Numeric |
| Explanatory (x) | Gestation           | Numeric |


## Insert Mine's UCI ICS 80 Sampling and Study Design (i.e. regression slides)

## Insert Mine's UCI STATS 115 Evaluating Regression Models
-->
